Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2021-08-12 write-up/covid-writeup.tex
Paper edits

\begin{abstract}
  Coronavirus case-count data has influenced government policies and drives most epidemiological forecasts. Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic. While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the COVID-19 pandemic; neither can be fully addressed by increased testing capacity. In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number. We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.  To address this, a procedure is presented that combines case-count data and random samples over time to estimate selection propensities based on key covariate information. We then combine these selection propensities with epidemiological forecast models using death-only data to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.  This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, the United States's only statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.  We end with a series of recommendations based on the proposed methodology.
\end{abstract}

@@

One core component of coronavirus research is epidemiological state-space modelling of case-count and death data. These models can be used to answer a variety of research questions including case-count forecasting, estimation of the effective reproduction number, and estimation of quarantine and other health policies on infectious disease dynamics.  The basic approach is a deterministic compartmental model called the susceptible-infectious-recovered (SIR) model.  A probabilistic extension was proposed by~\cite{Osthus2017} to model one-dimensional time series of infected proportions. \cite{Song2020} extended this approach to incorporate interventions and assess interventions on COVID-19 epidemic in China. \cite{Hao2020} builds an extended SEIR model (SAPHIRE) to account for various presymptomatic infectiousness, time-varying ascertainment rates, transmission rates and population movements. Given a probability sampling design, individual predictions can be leveraged to improve estimation via model-assisted approaches~\citep{Breidt2017}.  For nonprobability samples, \cite{Chen2019} derive a doubly robust approach that uses outcome predictions given covariates on the nonprobability and probability samples.  Here, we combine the state-space SIR approach of \cite{Song2020} but instead, as in \cite{Johndrow2020}, focus on COVID-19 confirmed death count data.  We generate epidemiological forecasts for point prevalence of each population strata.  We then demonstrate how to combine these forecasts with the IPW approach to construct doubly-robust estimates of point prevalence.  A derived statistical error decomposition guides this discussion.

\textcolor{red}{Discuss the additional biases and Qing paper.}

@@

The primary goal of this manuscript is to produce accurate estimates of the population-level active infection rates over time using \emph{publicly available} viral testing data.  Secondary goals include estimation of rates of change and the effective reproduction number which characterize disease trajectory.  A final goal is characterizing prevalence of population-level immunity. These quantities are fundamental to public health policy and provide critical information on the presence and transmission of SARS-CoV-2.
 % as well as the impact of public health  interventions.

@@

Coronavirus case-count data refers to the number of positive molecular tests performed on each day.  Figure~\ref{fig:in-cases} plots the number of reported confirmed COVID-19 cases per day in the state of Indiana.  Figure~\ref{fig:in-tests} plots the total number of COVID-19 molecularw tests performed per day. Figure~\ref{fig:in-deaths} plots the total number of COVID-19 related reported deaths per day. Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide accessible incoming county-level information of confirmed cases and deaths.

Public databases most often contain aggregate information.  The Johns Hopkins dashboard, for example, provides demographic breakdown of case counts as well as the total confirmed cases and deaths by county.  Aggregate time series of case count and deaths can also be extracted.  Unfortunately, most dashboards do not provide any demographic information on who requested a test nor on who tested positive for SARS-CoV-2 over time. Working closely with the State of Indiana, we were able to produce publicly available datasets for COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.  This granular dataset motivates the proposed approach.
% https://www.regenstrief.org/covid-dashboard/
%

@@

Due to testing restrictions and other potential selection biases, publicly reported Covid testing data may not be sufficient to understand active infection rates or the disease trajectory.  Here, we discuss two random samples that can be leveraged to as auxiliary information to address selection bias.

@@

\subsubsection{Facebook/CMU symptom survey.}
\label{subsection:fbsymptom}
Since April 2020, Facebook has started randomly showing an opt-in symptom survey to its users~\citep{delphisurvey}.  Data collected includes basic demographic information and if the respondent has symptoms such as fever, coughing, shortness of breath, or loss of smell which are associated with COVID-19.  The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.  Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals within age and gender strata who report a fever within the past 24-hours.  This paper uses the Facebook symptom survey as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., fraction of population displaying symptoms or in contact with COVID-19 positive individuals).

@@

To start, we assume the overall number of COVID-19 cases and/or disease prevalence are of primary interest. That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$. Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.  Then a natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.  Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.

@@

First we require some additional notation.  Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.  The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples.  In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity were known a priori, we present a novel iterative procedure that uses the error decomposition to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators. To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
\begin{equation}
\label{eq:statdecomp}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M},
\end{equation}
where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.  See Appendix~\ref{app:memestimator} for the derivation. This extends work by \cite{Meng2018} to account for imperfect testing. The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.  The final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence but here is scaled by the factor $(1 - FP - FN)^{-1}$.

Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}. Under either perfect testing (i.e., $FP=FN=0$), $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates. Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply scales the error. When tests are imperfect and selection bias exists, this simple relationship no longer holds.

@@

Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
$$
\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]
$$
implying the error relative to SRS increases as a function of population size~\citep{Meng2018}. The paradox that two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.  When comparing the US and UK under the assumption of similar testing strategies, for example, the MSE increases by a factor of five.  Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.

A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
$$
\frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
$$
As of June 10th 2020, the US has performed $21,474,531$ tests.  The US population is roughly $312$ million, so $f = 0.069$.  \textcolor{red}{FIX: The cumulative empirical prevalence is $9.3\%$ on May 8th. A recent study estimates a false positive and negative rate for antibody testing at $0.5$\% and $17.2$\% respectively \citep{Bendavid2020}. Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $92$.Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.  If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $1950$.  Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.  Moreover, increased testing capacity may also come at the cost of increases in false positive and negative rates.  Consider the case where $f$ increases from $0.05$ to $0.1$ and the increase is associated with the false negative rate rising from $5\%$ to $20$\%.  Then the effective sample size increases by a factor of $1.76$ rather than the expected $4$ fold increase.}
% \begin{wrapfigure}{r}{0.5\textwidth}

@@

\subsection{Computating the effective sample size}
\label{section:effss}

For binary outcomes, we have
\begin{equation} \label{eq:binaryrho}
\rho_{I,Y} = \Delta \sqrt{\frac{\bar Y (1 - \bar Y)}{f (1-f)} }
\end{equation}
where $\Delta = P_J (I_J = 1 \mid Y_J = 1) - P(I_J = 1 \mid Y_J = 0) = f_1 - f_0$.  Suppose that $M = f_1/f_0$; then $f_0 = f / (\bar Y \cdot (M-1) + 1)$.  Using the upper bound $E_{\I} [ \rho_{I,Y}^2 ] \leq E_{\I} [\rho_{I,Y} ]^2$, we compute effective sample size under a range of values $\bary$, $M$ with $f = 0.066$ (i.e., current sampling fraction) and observed prevalence $0.095$.

>>>>

\begin{abstract}
  Coronavirus case-count data has influenced government policies and drives most epidemiological forecasts. Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic. While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the COVID-19 pandemic; neither can be fully addressed by increased testing capacity. In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number. We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.  To address this, a procedure is presented that combines case-count data and random samples over time to estimate selection propensities based on key covariate information. We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.  This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, the United States's only statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.  We end with a series of recommendations based on the proposed methodology.
\end{abstract}

@@

One core component of coronavirus research is epidemiological state-space modelling of case-count and death data. These models can be used to answer a variety of research questions including case-count forecasting, estimation of the effective reproduction number, and estimation of quarantine and other health policies on infectious disease dynamics.  The basic approach is a deterministic compartmental model called the susceptible-infectious-recovered (SIR) model.  A probabilistic extension was proposed by~\cite{Osthus2017} to model one-dimensional time series of infected proportions. \cite{Song2020} extended this approach to incorporate interventions and assess interventions on COVID-19 epidemic in China. \cite{Hao2020} builds an extended SEIR model (SAPHIRE) to account for various presymptomatic infectiousness, time-varying ascertainment rates, transmission rates and population movements.

Recent work by \cite{Zhao2021} pointed out that estimation of key epidemiological parameters such as the incubation time using standard epidemiological models can suffer from severe bias due to issues beyond selection bias and measurement error.  Namely, right truncation and epidemic growth leading to patients ``being more likely to be infected towards the end of their exposure period''~\cite[pp. 3]{Zhao2021}.  Their approach carefully constructs the study sample and statistical model to account for these issues.  In this paper, rather than buildinga  study sample, we ask whether we can collect auxiliary information to address selection bias in the observed case count data.

Given a probability sampling design, individual predictions can be leveraged to improve estimation via model-assisted approaches~\citep{Breidt2017}.  For nonprobability samples, \cite{Chen2019} derive a doubly robust approach that uses outcome predictions given covariates on the nonprobability and probability samples.  Here, we combine the state-space SEIR approach of \cite{Song2020} but instead, as in \cite{Johndrow2020}, focus on COVID-19 confirmed death count data.  We generate epidemiological forecasts for point prevalence of each population strata.  We then demonstrate how to combine these forecasts with the IPW approach to construct doubly-robust estimates of point prevalence.  A derived statistical error decomposition guides this discussion.

@@

The primary goal of this manuscript is to produce accurate estimates of the population-level active infection rates over time using \emph{publicly available} viral testing data.  Secondary goals include estimation of rates of change and the effective reproduction number which characterize disease trajectory.
% A final goal is characterizing prevalence of population-level immunity.
These quantities are fundamental to public health policy and provide critical information on the presence and transmission of SARS-CoV-2.
 % as well as the impact of public health  interventions.

@@

Coronavirus case-count data refers to the number of positive molecular tests performed on each day.  Figure~\ref{fig:in-cases} plots the number of reported confirmed COVID-19 cases per day in the state of Indiana.  Figure~\ref{fig:in-tests} plots the total number of COVID-19 molecular tests performed per day. Figure~\ref{fig:in-deaths} plots the total number of COVID-19 related reported deaths per day. Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide accessible incoming county-level information of confirmed cases and deaths.

Public databases most often contain aggregate information.  The Johns Hopkins dashboard, for example, provides demographic breakdown of case counts as well as the total confirmed cases and deaths by county.  Aggregate time series of case count and deaths can also be extracted.  Unfortunately, most dashboards do not provide any demographic information on who requested a test nor on who tested positive for SARS-CoV-2 over time. Working closely with the State of Indiana, we were able to obtain access to COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.  These granular datasets are now publicly available and motivate the proposed approach.
% https://www.regenstrief.org/covid-dashboard/
%

@@

Due to testing restrictions and other potential selection biases, publicly reported COVID-19 testing data may not be sufficient to understand active infection rates or the disease trajectory.  Here, we discuss two random samples that provide auxiliary information and may help address selection bias.

@@

\subsubsection{Facebook/CMU symptom survey.}
\label{subsection:fbsymptom}
Since April 2020, Facebook has started randomly showing an opt-in symptom survey to its users~\citep{delphisurvey}.  Data collected includes basic demographic information and if the respondent has symptoms such as fever, coughing, shortness of breath, or loss of smell which are associated with COVID-19.  The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.  Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals within age and gender strata who report a fever within the past 24-hours.  This paper uses the Facebook symptom survey as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., fraction of population displaying symptoms or in contact with COVID-19 positive individuals.

@@

To start, we assume the overall number of active COVID-19 cases and/or active infection rate (AIR) are of primary interest. That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$. Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.  Then a natural candidate for AIR is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall active cases is $N \times \bar y$.
Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.  Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.

@@

First we require some additional notation.  Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.  The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples.  In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity are known a priori, a novel iterative procedure is used to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators. To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
\begin{equation}
\label{eq:statdecomp}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M},
\end{equation}
where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.  See Appendix~\ref{app:memestimator} for the derivation. This extends work by \cite{Meng2018} to account for imperfect testing. The first three terms continue to represent \emph{data quality}, \emph{data quantity}, and \emph{problem difficulty} respectively.  The new term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence, which is scaled by the factor $(1 - FP - FN)^{-1}$.

Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}. Under perfect testing (i.e., $FP=FN=0$), $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates. Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply scales the error. When tests are imperfect and selection bias exists, this simple relationship no longer holds.

@@

Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
$$
\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]
$$
implying the error relative to SRS increases as a function of population size~\citep{Meng2018}. The paradox that two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.  Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.

A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
$$
\frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
$$
Between April 25th to 29th 2020, Indiana performed $95,879$ tests.  Indiana's population is roughly $6.732$ million, so $f = 0.003$.  The active infection rate is $11.7\%$ in this time interval. Recent studies have suggested 87\% sensitivity~\cite{Arevalo2020} and 97.6\% specificity~\cite{Cohen2020} are reasonable measurement error estimates for RT-PCR molecular tests. Supposing COVID-19 positive individuals are $1.3$ times more likely to get tested, then the effective sample size is $84$. Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.  Even if the relative sampling rate drops to $1.1$ and $f$ increases to $0.01$ then the effective sample size will increase to $714$.  Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.  Moreover, increased testing capacity may also come at the cost of increases in false positive and negative rates.  Consider the case where $f$ increasing from $0.003$ to $0.01$ and is associated with the false negative rate rising from $13\%$ to $20$\%.  Then the effective sample size is $602$, representing a $7.2$ factor increase rather than the expected $8.5$ factor increase.
% \begin{wrapfigure}{r}{0.5\textwidth}

@@

\subsection{Computing the effective sample size}
\label{section:effss}

For binary outcomes, we have
\begin{equation} \label{eq:binaryrho}
\rho_{I,Y} = \Delta \sqrt{\frac{\bar Y (1 - \bar Y)}{f (1-f)} }
\end{equation}
where $\Delta = P_J (I_J = 1 \mid Y_J = 1) - P(I_J = 1 \mid Y_J = 0) = f_1 - f_0$.  Suppose that $M = f_1/f_0$; then $f_0 = f / (\bar Y \cdot (M-1) + 1)$.  Using the upper bound $E_{\I} [ \rho_{I,Y}^2 ] \leq E_{\I} [\rho_{I,Y} ]^2$, we compute effective sample size under a range of prevalences $\bary$, and relative sample rates $M$ given $f = 0.003$ (i.e., current sampling fraction).
